Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-05-18 second-draft/pnas-word-order-si.tex
SI

\textcolor{red}{TODOs}

\textcolor{red}{TODO This idea makes strong predictions, and turns out to overpredict correlations (see SI, Section X).}

@@

@@

\paragraph{Per-Language Results}
Figure~\ref{fig:pareto-per-lang}

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage.pdf}
\caption{per language results. Green: random baselines, Red: real grammar, blue: approximate Pareto frontier, computed from the optimized grammars.\textcolor{red}{would be good to have same colors as in main paper} All data are $z$-transformed.}\label{fig:pareto-per-lang}
\end{figure}

@@

Further, we report results for grammars optimized for dependency length minimization (DLM).
Following \cite{futrell2015largescale}, we formalize the dependency length of a sentence as the sum of the distances of all pairs of syntactically related words, where distance is measured by the number of intervening words.
Prior work has suggested principles akin to DLM as approximating efficiency optimization of grammars~\cite{hawkins1994performance,futrell2017memory, futrell2017generalizing}.
Indeed, we found that DLM is predicted by efficiency optimization, as reported in Discussion (Main Paper).
In line with these ideas, we show that optimizing for DLM makes predictions similar to efficiency optimization, itself a novel result.

@@

&	Relation & Real & DepL & Pred & Pars & Efficiency & Expected Prevalence  \\ \hline

@@

\begin{table*} % TODO figure out away to make this table more aesthetic
	\begin{center}	
\small{
\begin{tabular}{|l|l|l|ll|l|l|llllllllllllllllllllllllllll}
	\hline
	Relation & Real & DepL & Pred & Pars & Efficiency & Expected Prevalence  \\ \hline
\hline
appos  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_appos.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_appos.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_appos.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_appos.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_appos.pdf}   &                                               Unknown \\           
lifted\_cc  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_lifted_cc.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_lifted_cc.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_lifted_cc.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_lifted_cc.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_lifted_cc.pdf}   &                  Unknown \\          
expl  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_expl.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_expl.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_expl.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_expl.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_expl.pdf}   &                                                        Unknown \\         
iobj  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_iobj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_iobj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_iobj.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_iobj.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_iobj.pdf}   &                                                        Unknown \\     
vocative  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_vocative.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_vocative.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_vocative.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_vocative.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_vocative.pdf}   &                            Unknown \\

@@

Greater values of $\lambda$ mathematically result in degenerate solutions.
To show this, note that the following inequality always holds:
\begin{equation}
I[L(T); T] \leq H[L(T)].
\end{equation}
Therefore, if $\lambda \geq 1$, the efficiency objective satisfies 
\begin{equation}
    J_T(L) = I[L(T); T] - \lambda H[L(T)] \leq 0.
\end{equation}
and it takes the maximally possible value zero if $L(T)$ maps all trees to a single utterance, in which case both $I[L(T); T]$ and $H[L(T)]$ are zero.
This is a degenerate language with only a single utterance, which is simultaneously used to convey all meanings.
While the design of our word order grammars precludes a collapse of all trees to a single utterance, this shows that an objective with $\lambda \geq 1$ cannot be a generally applicable objective for language efficiency.

In our experiments, we chose $\lambda = 0.9$ as a mathematically valid value that puts similar weight on both predictability and parsability.

The computational cost of grammar optimization precluded repeating the experiment for many values of $\lambda$.
We previously carried out language optimization for $\lambda = 1$, and obtained essentially identical predictions for word order correlations.
We note that each of the eight correlations is predicted by at least parseability or predictability, without any contradictory predictions (Table~\ref{tab:all-predictions-1}).

@@

In this section, we describe the details of the neural network architectures.
All choices described in this section were fixed before evaluating word order properties and the efficiency of real language.

@@

\paragraph{Estimating Predictability}
We choose a standard LSTM \citep{hochreiter1997long} language model with vocabulary size restricted to the most frequent 50,000 words in the treebanks for a given language.
Given the small size of the corpora, this limit is only attained only for few languages.
In each time step, the input is a concatenation of embeddings for the word, for language-specific POS tags, and for universal POS tags.
The model predicts both the next word and its POS tags in each step.
Using POS tags is intended to prevent overfitting on small corpora.
These decisions were made before evaluating word order properties.

@@

Neural network models have hyperparameters such as the number of hidden units, and the learning rate. 
For predictability and parseability optimization, we first selected hyperparameters on the respective objectives for selected languages on the provided development partitions.
Then, for each language and each objective function, we created eight random combinations of these selected hyperparameter values, and selected the setting that yielded the best value of the respective objective function (predictability or parseability) on the language. We then used this setting for creating optimized word order grammars.

\textcolor{red}{TODO random seeds for efficiency}

@@

%\begin{table}[]
%    \centering
%    \begin{tabular}{c|cccc}
%    Entropy penalty     &  0.0001, 0.001 \\
%    Learning Rate (Word order grammar)     & 5e-6, 1e-5, 2e-5, 5e-5 \\
%    Momentum (Word order grammar) & 0.8, 0,9 \\
%    Dropout (Language model) & 0.0, 0.3, 0.5 \\
%    
%    \end{tabular}
%    \caption{Hyperparameters}
%    \label{tab:my_label}
%\end{table}

\paragraph{Optimization details}

We employ two common variance reduction methods to improve the estimator~(\ref{eq:dtheta}), while keeping it unbiased.
%For dependency length and 
For predictability, note that the loss incurred for a specific word only depends on ordering decisions made up to that word (and its head, in the case of dependency length). We represent the process of linearizing a tree as a dynamic stochastic computation graph, and use these independence properties to apply the method described in \citet{schulman2015gradient} to obtain a version of~(\ref{eq:dtheta}) with lower variance.
Second, we use a word-dependent moving average of recent per-word losses as control variate \cite{williams1992simple}.
These two methods reduce the variance of the estimator and thereby increase the speed of optimization and reduce training time, without biasing the results.
For numerical stability, we represent $a_\tau \in [0,1]$ via its logit $\in \mathbb{R}$.
Furthermore, to encourage exploration of the parameter space, we add an entropy regularization term \citep{xu2015show} for each Direction Parameter $a_\tau$, which penalizes $a_\tau$ values near $0$ or $1$. The weight of the entropy regularization was chosen together with the other hyperparameters.
%, adding the negative entropy of the $Bernoulli(a_\tau)$ to the loss $R$. %, weighted with $\alpha$.

@@

We update word order grammar parameters $\theta$ using Stochastic Gradient Descent with momentum.
For the language model parameters $\phi$, we use plain SGD. 
For the parser parameters $\psi$, we use Adam \citep{kingma2014adam}, following \citet{dozat2017stanford}.
The learning rates and other optimization hyperparameters were determined together with the other hyperparameters.

For each language and objective function, we apply Early Stopping~\citep{prechelt1998early} as follows. We compute a Monte-Carlo estimate of the objective on the UD development set in intervals of 50,000 training steps, and stop optimization once the value of the objective deteriorates compared to the last previous estimate. This technique, well known in the machine learning literature, helps prevent overfitting to the specific corpora used.

@@

\paragraph{Optimized Grammars}

Recall that, for each language, we created 8 optimized languages for each optimization criterion.
We enforced balanced distribution of object-verb and verb-object ordering among optimized languages by fixing $a_\tau$ for the \textit{obj} dependency to be 0.0 in four of these languages, and 1.0 in the other four.
This maximizes statistical precision in detecting and quantifying correlations between the \textit{obj} relation and other relations.

For efficiency optimization, for each grammar, we ran efficiency optimization with three different random seeds, selecting among these the seed that yielded the best overall efficiency value.
This

@@

\section{Dependency Length Minimization}

@@

>>>>

@@

%\textcolor{blue}{Greenberg~\cite{greenberg1963universals} stated a total of 45 universals.
%Twenty of these concern the structure of individual words (as opposed to word order, which we focus on here), and many of those have been argued to be explained by the ``dual pressures'' idea \cite{haspelmath2006against}.
%The other 25 universals concern word order; Dryer \cite{dryer1992greenbergian} reformulated most of these as correlations with verb-object order, finding 16 correlations to likely be universal.
%Excluding universals not annotated in available databases, and merging some that can only be tested together we arrived at the eight correlations shown in Table~\ref{table:corr-dryer}.
%Testing efficiency on the excluded correlations would be feasible with additional annotation of data.
%For instance, one of these could be tested by labeling the existing databases for the distinction between manner adverbs from other kinds of adverbs.
%For another one, it would be necessary to annotate databases for additional languages that have independent plural markers (e.g., Hawaiian).}

@@

\paragraph{Per-Language Results}
In Figure~\ref{fig:pareto-per-lang}, we show the predictability-parseability planes for every one the 51 languages, together with Pareto frontiers estimated from optimized grammars.\footnote{Note that in a few languages, the real grammar is at a position slightly \emph{beyond} the estimated Pareto frontier. Due to use of stochastic gradient descent in optimization, there is some amount of noise in the optimization process.}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage.pdf}
\caption{Predictability and parseability of 51 languages. Green: random baselines, Red: real grammar, blue: approximate Pareto frontier, computed from the optimized grammars. All data are $z$-scored.}\label{fig:pareto-per-lang}
\end{figure}

@@

Further, we report results for grammars optimized for dependency length minimization (DLM, see Section~\ref{sec:DLM}).
Following \cite{futrell2015largescale}, we formalize the dependency length of a sentence as the sum of the distances of all pairs of syntactically related words, where distance is measured by the number of intervening words.
We optimize grammars using the same stochastic gradient descent method as we use for efficiency optimization (Section REF).
Prior work has suggested principles akin to DLM as approximating efficiency optimization of grammars~\cite{hawkins1994performance,futrell2017memory, futrell2017generalizing}.
Indeed, we found that DLM is predicted by efficiency optimization, as discussed in Section REF.
Results in Table~\ref{tab:all-predictions-1} show that optimizing for DLM makes predictions similar to efficiency optimization, itself a novel result.

@@

&	Relation & Real & DLM & Pred & Pars & Efficiency & Expected Prevalence  \\ \hline

@@

\begin{table*} % TODO figure out away to make this table more aesthetic
	\begin{center}	
\small{
\begin{tabular}{|l|l|l|ll|l|l|llllllllllllllllllllllllllll}
	\hline
	Relation & Real & DLM & Pred & Pars & Efficiency & Expected Prevalence  \\ \hline
\hline
appos  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_appos.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_appos.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_appos.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_appos.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_appos.pdf}   &                                               Unknown \\           
lifted\_cc  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_lifted_cc.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_lifted_cc.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_lifted_cc.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_lifted_cc.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_lifted_cc.pdf}   &                  Unknown \\          
expl  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_expl.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_expl.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_expl.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_expl.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_expl.pdf}   &                                                        Unknown \\         
iobj  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_iobj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_iobj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_iobj.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_iobj.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_iobj.pdf}   &                                                        Unknown \\     
vocative  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_vocative.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_vocative.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_vocative.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_vocative.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_vocative.pdf}   &                            Unknown \\

@@

Greater values of $\lambda$ mathematically result in degenerate solutions.
To show this, note that the following inequality always holds:
\begin{equation}
I[L(T); T] \leq H[L(T)].
\end{equation}
Therefore, if $\lambda \geq 1$, the efficiency objective satisfies 
\begin{equation}
    J_T(L) = I[L(T); T] - \lambda H[L(T)] \leq 0.
\end{equation}
and it takes the maximally possible value zero if $L(T)$ maps all trees to a single utterance, in which case both $I[L(T); T]$ and $H[L(T)]$ are zero.
This is a degenerate language with only a single utterance, which is simultaneously used to convey all meanings.
While the design of our word order grammars precludes a collapse of all trees to a single utterance, this shows that an objective with $\lambda \geq 1$ cannot be a generally applicable objective for language efficiency.
In conclusion, $\lambda$ is constrained to be in $[0,1)$, which values closer to $1$ placing similar weights on both predictability and parsability, whereas values closer to $0$ diminish the role of predictability.

In our experiments, we chose $\lambda = 0.9$ as a mathematically valid value that puts similar weight on both predictability and parsability.
The computational cost of grammar optimization precluded repeating the experiment for many values of $\lambda$.
%We previously carried out language optimization for $\lambda = 1$, and obtained essentially identical predictions for word order correlations, see Table REF.
We note that each of the eight correlations is predicted by at least parseability or predictability, without any contradictory predictions (Table~\ref{tab:all-predictions-1}).
%Indeed, the same holds for all of the 38 syntactic relations in UD  (Table~\ref{tab:all-predictions-1}-\ref{tab:all-predictions-2})
%We additionally report results with $\lambda = 0$, i.e., placing all weight on parseability.
That is, at $\lambda$ close to its maximal value, the predictions of optimizing the two scoring functions individually add up to the predictions of efficiency optimization.
Small values of $\lambda$ correspond to the case where predictability plays no role, and only parseability is optimized (Table~\ref{tab:all-predictions-1}).
We also verified the robustness of our results by reporting results with $\lambda=1$ in Table REF>

@@

In this section, we describe the details of the neural network architectures.
Choices follow standard practice and machine learning and were fixed before evaluating word order properties, and the efficiency of real languages.

@@

\paragraph{Estimating Predictability}
We choose a standard LSTM \citep{hochreiter1997long} language model with vocabulary size restricted to the most frequent 50,000 words in the treebanks for a given language.
Given the small size of the corpora, this limit is only attained only for few languages.
In each time step, the input is a concatenation of embeddings for the word, for language-specific POS tags, and for universal POS tags.
The model predicts both the next word and its POS tags in each step.
Using POS tags is intended to prevent overfitting on small corpora.

@@

Neural network models have hyperparameters such as the number of hidden units, and the learning rate. 
For predictability and parseability optimization, we first selected hyperparameters on the respective objectives for selected languages on the provided development partitions.
These parameters are shown in Table~\ref{tab:hyperparameters}.
Then, for each language and each objective function, we created eight random combinations of these selected hyperparameter values, and selected the setting that yielded the best value of the respective objective function (efficiency, predictability, parseability) on the language. We then used this setting for creating optimized word order grammars.

%\textcolor{red}{TODO random seeds for efficiency}

@@

\begin{table}[]
    \centering
    \begin{tabular}{c|cccc}
    Entropy penalty     &  0.0001, 0.001 \\
    Learning Rate (Word order grammar)     & 5e-6, 1e-5, 2e-5, 5e-5 \\
    Momentum (Word order grammar) & 0.8, 0,9 \\
    Dropout (Language model) & 0.0, 0.3, 0.5 \\

    \end{tabular}
    \caption{Selected Hyperparameters}
    \label{tab:hyperparameters}
\end{table}

\paragraph{Improved Unbiased Gradient Estimator}
We employ two common variance reduction methods to improve the estimator~(\ref{eq:dtheta}), while keeping it unbiased.
%For dependency length and 
% (and its head, in the case of dependency length)
For predictability, note that the loss incurred for a specific word only depends on ordering decisions made up to that word.
We represent the process of linearizing a tree as a dynamic stochastic computation graph, and use these independence properties to apply the method described in \citet{schulman2015gradient} to obtain a version of~(\ref{eq:dtheta}) with lower variance.
Second, we use a word-dependent moving average of recent per-word losses as control variate \cite{williams1992simple} (the word's surprisal in the case of predictability, and the negative log-probability of the correct head and relation label in the case of parseability).
These two methods reduce the variance of the estimator and thereby increase the speed of optimization and reduce training time, without biasing the results.
For numerical stability, we represent $a_\tau \in [0,1]$ via its logit $\in \mathbb{R}$.
Furthermore, to encourage exploration of the parameter space, we add an entropy regularization term \citep{xu2015show} for each Direction Parameter $a_\tau$, which penalizes $a_\tau$ values near $0$ or $1$. The weight of the entropy regularization was chosen together with the other hyperparameters.
%, adding the negative entropy of the $Bernoulli(a_\tau)$ to the loss $R$. %, weighted with $\alpha$.

@@

\paragraph{Learning Rate and Stopping Criterion}
We update word order grammar parameters $\theta$ using Stochastic Gradient Descent with momentum.
For the language model parameters $\phi$, we use plain Stochastic Gradient Descent without momentum, as recommended by \citet{merity2018regularizing}. 
For the parser parameters $\psi$, we use Adam \citep{kingma2014adam}, following \citet{dozat2017stanford}.
The learning rates and other optimization hyperparameters were determined together with the other hyperparameters.

For gramar optimization, we apply Early Stopping~\citep{prechelt1998early} as follows. We compute a Monte-Carlo estimate of the objective on the UD development set in intervals of 50,000 training steps, and stop optimization once the value of the objective deteriorates compared to the last previous estimate.
This is a standard technique for optimizing neural models, and helps prevent overfitting to the specific corpora used.

@@

\paragraph{Optimized Grammars}
Recall that, for each language, we created 8 optimized languages for each optimization criterion.
We enforced balanced distribution of object-verb and verb-object ordering among optimized languages by fixing $a_\tau$ for the \textit{obj} dependency to be 0.0 in four of these languages, and 1.0 in the other four.
This maximizes statistical precision in detecting and quantifying correlations between the \textit{obj} relation and other relations.

For efficiency optimization, for each grammar, we ran efficiency optimization with three different random seeds, selecting among these the seed that yielded the best overall efficiency value.
This helps control for variation across random seeds for the stochastic gradient descent optimization method.

@@

\section{Dependency Length Minimization}\label{sec:DLM}

@@

Further, we report results for grammars optimized for dependency length minimization (DLM, see Section~\ref{sec:DLM}).
Following \cite{futrell2015largescale}, we formalize the dependency length of a sentence as the sum of the distances of all pairs of syntactically related words, where distance is measured by the number of intervening words.
Prior work has suggested principles akin to DLM as approximating efficiency optimization of grammars~\cite{hawkins1994performance,futrell2017memory, futrell2017generalizing}.
Indeed, we found that DLM is predicted by efficiency optimization, as reported in Discussion (Main Paper).
In line with these ideas, we show that optimizing for DLM makes predictions similar to efficiency optimization, itself a novel result.
